---
title: "Accessing Jornada data in EDI"
author: "Greg Maurer, Darren James"
date: '2026-07-31'
execute:
  eval: false
# output:
#     html_document:
#       toc: true
---

```{r}
#| label = "setup",
#| include = FALSE
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This lesson will teach you several methods for accessing Jornada data at the
[Environmental Data Initiative (EDI) repository](https://edirepository.org). 
EDI has recently become more strict about data access and now requires authentication
for downloading data files.

## 1. Downloading from the portal

The EDI Portal ([portal.edirepository.org](https://portal.edirepository.org)) is the
main graphical interface to the repository. It provides a search interface and landing
pages for all datasets, and provides a way to download the files published as a part
of any dataset.

Downloading data requires "authentication" - i.e. you have to be logged in with a
non-anonymous account. EDI makes it easy to use your pre-existing Google or GitHub account
to authenticate.

## 2. EDIutils

The [EDIutils](https://docs.ropensci.org/EDIutils/) R package is a wrapper around the EDI
repository's API. It provides functions to search and access the data published in 
EDI from the comfort of your R scripts

```{r}
#| message = FALSE
library(tidyverse)
```

Then, load a comma-separated value file from the EDI repository. At the EDI
repository, each file in a dataset is assigned a download URL. This can be 
passed to the `tidyverse::read_csv()` function to read that file into a
dataframe.

```{r}
#| label = "load_ds1"
# Assign the address for the csv file hosted on EDI to a variable
infile <- "https://pasta.lternet.edu/package/data/eml/knb-lter-jrn/210011003/105/127124b0f04a1c71f34148e3d40a5c72"

# Read this file into a tibble with `tidyverse::read_csv()`
anpp.annual <- read_csv(infile, na=c('NA', '', '.'))
```

R tells us that we have two character data columns (`zone` and `site`), and two
floating point (`dbl`) columns that contain the year of measurement - `year` - 
and estimated annual NPP values (`npp_g_m2`). Lets make a simple plot to look at
the data. To do this we are using the `ggplot()` function, which is part of the 
`ggplot2` library that was attached when we loaded the `tidyverse` package`

```{r}
ggplot(anpp.annual, aes(x = year, y = npp_g_m2, col = site, group = site)) +
  geom_line() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

Now, lets output this into our `data/` directory so that we can load it easily
next time.

```{r}
write_csv(anpp.annual, '../data/td01_anpp.annual.csv')
```
