---
title: "JEST Introduction and Setup"
author: "Greg Maurer, Darren James"
date: '2022-06-18'
output:
    html_document:
      toc: true
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

Welcome to the Jornada Ecological Statistics Tutorials, or JEST! Our aim is to
teach introductory and intermediate statistics to Jornada researchers using R,
and real data from the Jornada.

## Setting up to run a JEST tutorial

We use the [R programming language](https://www.r-project.org/), and additional
packages from the R software ecosystem, as our primary statistical analysis
tools in these tutorials. During JEST workshops we teach using the [RStudio development environment](https://www.rstudio.com/) as our primary programming and analysis
environment. For working through these tutorials, either on your own or in a
workshop, you should have both installed according to the instructions below.

*The section below borrows heavily from the "[Data Carpentry: Data Analysis and
Visualization in R for Ecologists](https://datacarpentry.org/R-ecology-lesson/)"
lesson (Eunice, 2022)*

```{r results="asis", echo=FALSE}
# This brings in DC R ecology content for installing R and R studio
dc_lines <- readLines('https://github.com/datacarpentry/R-ecology-lesson/raw/main/index.Rmd')
cat(dc_lines[55:140], sep = '\n')
```


### Install required R packages

During the course we will need a number of R packages. Packages contain useful 
R code written by other people. We will use the packages 
`tidyverse`, `nlme`, `lme4`, `emmeans`, and `AICmodavg`.

To try to install these packages, open RStudio and copy and paste the following 
command into the console window (look for a blinking cursor on the bottom left),
then press the <kbd>Enter</kbd> (Windows and Linux) or <kbd>Return</kbd> (MacOS) 
to execute the command.

```{r eval=FALSE}
install.packages(c("tidyverse", "nlme", "lme4", "emmeans", "AICmodavg"))
```

Alternatively, you can install the packages using RStudio's graphical user 
interface by going to `Tools > Install Packages` and typing the names of the 
packages separated by a comma.

R tries to download and install the packages on your machine.
When the installation has finished, you can try to load the
packages by pasting the following code into the console:

```{r eval=FALSE}
library(tidyverse)
library(nlme)
library(lme4)
library(emmeans)
library(AICmodavg)
```

If you do not see an error like `there is no package called ‘...’` you are good 
to go!

```{r results="asis", echo=FALSE}
# This brings in DC R ecology content for updating R packages
#dc_lines <- readLines('https://github.com/datacarpentry/R-ecology-lesson/raw/main/index.Rmd')
cat(dc_lines[174:194], sep = '\n')
```

The data files for each tutorial will be specified near the top of the page.

## Preparing for JEST workshops

These workshops are based on "live coding," so instructors will be running
example R code while learners are expected to follow along on their own machine.
To follow along successfully during a JEST workshop, please ensure you have a
laptop or other Windows, MacOS, or Linux device that has R and RStudio, and any
necessary R packages, installed and up to date. Also bring a charger for your
device.

If the workshop is in a virtual format we will be using the Zoom platform, so
make sure you have Zoom installed and up-to-date on your machine as well.

