---
title: "JEDS Introduction and Setup"
author: "Greg Maurer, Darren James"
date: '2022-06-18'
output:
    html_document:
      toc: true
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Welcome to the Jornada Environmental Data Science education site (JEDS!) Our aim is 
to teach introductory and intermediate elements of data science (data management, 
analysis, statistics, visualization, etc.) to Jornada researchers using open-source
software tools, and real data from the Jornada.

## Setting up for an `R` lesson

Many JEDS tutorials use the [R programming language](https://www.r-project.org/), and additional
packages from the R software ecosystem, as the primary software tools, especially for statistical analysis.
During `R`-based JEDS workshops we teach using the [RStudio development environment](https://www.rstudio.com/)
as our primary programming and analysis environment. For working through these tutorials, either on your own
or in a workshop, you should have both installed according to the instructions below.

*The section below borrows heavily from the "[Data Carpentry: Data Analysis and
Visualization in R for Ecologists](https://datacarpentry.org/R-ecology-lesson/)"
lesson (Eunice, 2022)*

```{r results="asis", echo=FALSE}
# This brings in DC R ecology content for installing R and R studio
dc_lines <- readLines('https://github.com/datacarpentry/R-ecology-lesson/raw/main/index.Rmd')
cat(dc_lines[55:140], sep = '\n')
```


### Install required `R` packages

During the course we will need a number of R packages. Packages contain useful 
R code written by other people. In most workshops will use the packages 
`tidyverse`, `nlme`, `lme4`, and `emmeans`.

To try to install these packages, open RStudio and copy and paste the following 
command into the console window (look for a blinking cursor on the bottom left),
then press the <kbd>Enter</kbd> (Windows and Linux) or <kbd>Return</kbd> (MacOS) 
to execute the command.

```{r eval=FALSE}
install.packages(c("tidyverse", "nlme", "lme4", "emmeans"))
```

Alternatively, you can install the packages using RStudio's graphical user 
interface by going to `Tools > Install Packages` and typing the names of the 
packages separated by a comma.

R tries to download and install the packages on your machine.
When the installation has finished, you can try to load the
packages by pasting the following code into the console:

```{r eval=FALSE}
library(tidyverse)
library(nlme)
library(lme4)
library(emmeans)
```

If you do not see an error like `there is no package called ‘...’` you are good 
to go!

**NOTE** that if you are attending a workshop, you will likely receive additional
instructions about packages required for for your workshop. It is a good idea to
install these in advance, using commands like those above, and let your
instructors know if you have any problems.

```{r results="asis", echo=FALSE}
# This brings in DC R ecology content for updating R packages
#dc_lines <- readLines('https://github.com/datacarpentry/R-ecology-lesson/raw/main/index.Rmd')
cat(dc_lines[174:194], sep = '\n')
```

A copy of all data files used in these lessons are stored in the `data/`
directory of the GitHub repository. The data files you will work with will
be specified at the start of the lesson materials, and you may also receive
additional instructions or materials for workshops.

## Setup for shell, git and GitHub lessons

[`git`](https://git-scm.com/) is a distributed version control system and is the
underlying technology behind [GitHub](https://github.com), the web-based software development and
collaboration platform. It is common to use `git` together with a "shell", which
is a command-line interface to your computer. So, for JEDS tutorials that involve
`git` and GitHub, please complete three steps using the recommended resources:

1. Install `git` to your computer
2. Make sure you have access to a shell
3. Sign up for a [GitHub](https://github.com) account

If you are on MacOS or Linux, you should have a shell installed already, just make
sure you can open it. For Windows you can install `git` and the `bash` shell together.
There are two good resources for getting `git` and related software ready for a
one of the JEDS tutorials. 

  * The [Carpentries `git` installation instructions](https://carpentries.github.io/workshop-template/#git)
    gives good platform-based instructions.
  * [Happy Git with R](https://happygitwithr.com/install-intro.html) has more detailed instructions
    that cover a few edge-cases.

If you want, you can also install a GUI client for `git` to help visualize
and manage what is happening with repositories you are working with. A few recommended
clients are [GitKraken](https://www.gitkraken.com/),
[SourceTree](https://www.sourcetreeapp.com/), and [GitHub Desktop](https://desktop.github.com/).
More discussion of clients is available over at [Happy Git with R](https://happygitwithr.com/git-client.html).

```{r results="asis", include=FALSE, echo=FALSE}
# This brings in SC Shell/Git content for installing git
sc_lines <- readLines('https://github.com/carpentries/workshop-template/raw/gh-pages/_includes/install_instructions/git.html')
cat(sc_lines, sep = '\n')
```

## Preparing for JEDS workshops

These workshops are based on "live coding," so instructors will be running
example R code while learners are expected to follow along on their own machine.
To follow along successfully during a JEDS workshop, please ensure you have a
laptop or other Windows, MacOS, or Linux device that has R and RStudio, and any
necessary R packages, installed and up to date. Look for instructions provided
by instructors in advance of the workshop about what R packages to install, and
any other materials you may need to download or bring. Also, always bring a
charger for your device.

If the workshop is in a virtual format we will be using the Zoom platform, so
make sure you have Zoom installed and up-to-date on your machine as well.

